A new simplified manual tour, with examples in mathematica and R

### 2022-06-14 paper/paper.Rmd
fixed lists into Rmarkdown

The following functions are implemented in the Mathematica package {\tt mmtour.wl}

\begin{itemize}

\item ProjectionPlot[data, projmat, xRange(=Automatic), yRange(=Automatic), 
colour(=Automatic)] with arguments

\begin{itemize}
\item data: A list of data matrices (they must all have the same dimensionality)
\item projmat: A projection matrix which describes the projection plane. This is the 
initial projection and does not neet by orthonormal.
\item xRange: A list which details the range of the horizontal axis.
\item yRange: A list which details the range of the vertical axis.
\item colour: A list of graphics directives to be applied to each data \
matrix.
\end{itemize}

@@

\item SlicePlot[data, projmat, centrePoint, height, heightRange, 
ptSize1(=0.005), ptSize2(=0.004), minDist(=0)] with arguments (used to be called SliceDynamic in Alex's first version, the name SliceDynamic was taken by something else below)

\begin{itemize}
\item data: A data matrix
\item projmat: The inital projection matrix
\item centrepoint: The initial position of the slice
\item height: The initial height of the slice
\item heightRange: The range of heights
\item ptSize1: The size of the points that exist within the slice
\item ptSize2: The size of the points that exist outside of the slice
\end{itemize}

@@

\item SliceDynamic[data, projmat, centrePoint, height, heightRange, 
 ptSize1 : 0.005, ptSize2 : 0.004] with arguments

\begin{itemize}
\item data: A data matrix with an extra groupling or clustering column (numerical) 
\item projmat: The inital projection matrix
\item centrepoint: The initial position of the slice
\item height: The initial height of the slice
\item heightRange: The range of heights
\item ptSize1: The size of the points that exist within the slice
\item ptSize2: The size of the points that exist outside of the slice
\end{itemize}

@@

\item Projected2DSliderPlot[]

@@

\end{itemize}

>>>>

The following functions are implemented in the Mathematica package `mmtour.wl`:

- `ProjectionPlot[data, projmat, xRange(=Automatic), yRange(=Automatic), 
colour(=Automatic)]` with arguments

    - `data`: A list of data matrices (they must all have the same dimensionality)
    - ``projmat`: A projection matrix which describes the projection plane. This is the 
initial projection and does not neet by orthonormal.
    - ``xRange`: A list which details the range of the horizontal axis.
    - ``yRange`: A list which details the range of the vertical axis.
    - ``colour`: A list of graphics directives to be applied to each data \
matrix.

@@

- `SlicePlot[data, projmat, centrePoint, height, heightRange, ptSize1(=0.005), ptSize2(=0.004), minDist(=0)]` with arguments (used to be called SliceDynamic in Alex's first version, the name SliceDynamic was taken by something else below)

    - `data`: A data matrix
    - `projmat`: The inital projection matrix
    - `centrepoint`: The initial position of the slice
    - `height`: The initial height of the slice
    - `heightRange`: The range of heights
    - `ptSize1`: The size of the points that exist within the slice
    - `ptSize2`: The size of the points that exist outside of the slice

@@

- `SliceDynamic[data, projmat, centrePoint, height, heightRange, ptSize1 : 0.005, ptSize2 : 0.004]` with arguments

- `data`: A data matrix with an extra groupling or clustering column (numerical) 
- `projmat`: The inital projection matrix
- `centrepoint`: The initial position of the slice
- `height`: The initial height of the slice
- `heightRange`: The range of heights
- `ptSize1`: The size of the points that exist within the slice
- `ptSize2`: The size of the points that exist outside of the slice

@@

- `Projected2DSliderPlot[]`

@@

### 2022-06-14 paper/paper.tex
fixed lists into Rmarkdown

% see https://stackoverflow.com/a/47122900

% Pandoc citation processing

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\usepackage{graphicx}

@@

@@

@@

@@

>>>>

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{nicematrix}
\NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}

@@

\hypertarget{algorithm-3}{%
\subsubsection{Algorithm 3}\label{algorithm-3}}

For \(d=2\) projections, the projection matrix is the sub-matrix of
\(A\) formed by its first two columns. Whereas orthonormality of the
basis for the p-dimensional space is given by
\(e_i\cdot e_j=\delta_{ij},{i,j,=1,\cdots,}\), orthonormality of the
projection matrix is expressed as
\(P_i\cdot P_j=\delta_{ij}, {i,j=1,2}\). Movement of the cursor takes
the two components \({x_{m1},x_{m2}}\) into a selected new value
\({a,b}\). Although the motion is constrained by \(a^2+b^2\leq 1\), this
is not sufficient to guarantee orthonormality of the new projection
matrix. One possible algorithm to achieve this is

\begin{enumerate}
\item Cursor movement takes ${x_{m1},x_{m2}}\to {a,b}$
\item The freedom to change the components $A_{i3}\cdots A_{ip}$ (the columns of $A$ not corresponding to the projection matrix) is used to select a new orthonormal basis as follows:
\begin{enumerate}
\item For row $m$ one chooses $A_{m3}=\sqrt{1-a^2-b^2},~A_{m,k>3}=0$
\item For other rows, $A_{i\neq m,j\geq3}$ random selections in the range $(-1,1)$ are made.
\item The Gram-Schmidt algorithm is then used to obtain an orthonormal basis taking $e_m$ as the first vector (which is already normalised), and then proceeding as usual $e_1\to e_1-(e_1\cdot e_m) e_m$, $e_1\to e_1/(e_1\cdot e_1)$, etc.
\end{enumerate}
\item this results in the orthonormal basis $A^*$ and a new projection matrix with $P_{m1}=a,~P_{m2}=b$.
\end{enumerate}

The random choice for the components of \(A\) not in the projection
matrix allows the exploration of dimensions perpendicular to the
projection plane.

\begin{align}
A=~~~& \begin{pNiceArray}{ccccc}[first-row,first-col]
\CodeBefore
       \columncolor{red!15}{1,2}
              \rowcolor{blue!15}{4}
     \Body
       & \stackrel{P_1}{\downarrow} & \stackrel{P_2}{\downarrow} &  & &\\
e_1\to & x_{11} & x_{12}  & x_{13} & \cdots & x_{1p} \\
e_2 \to& x_{21}  & x_{22}  & \cdot  &\cdots & x_{2p} \\
\vdots ~~~&\vdots & \vdots  & \vdots & \ddots & \vdots \\ 
 e_m \to& x_{m1}  & x_{m2}  & \cdot  &\cdots & x_{mp} \\   
\vdots ~~~&\vdots & \vdots  & \vdots & \ddots & \vdots \\
e_p \to&x_{p1} & x_{p2} &\cdot &\cdots & x_{pp} 
\end{pNiceArray},\quad (x_{m1},x_{m2})\to (a,b)
\end{align}

@@

XXX need to check calculation of angle when \(m_1 < 0\)

@@

The following functions are implemented in the Mathematica package
\texttt{mmtour.wl}:

\begin{itemize}
\item
  \texttt{ProjectionPlot{[}data,\ projmat,\ xRange(=Automatic),\ yRange(=Automatic),\ \ colour(=Automatic){]}}
  with arguments

  \begin{itemize}
  \tightlist
  \item
    \texttt{data}: A list of data matrices (they must all have the same
    dimensionality)
  \item
    `\texttt{projmat}: A projection matrix which describes the
    projection plane. This is the initial projection and does not neet
    by orthonormal.
  \item
    `\texttt{xRange}: A list which details the range of the horizontal
    axis.
  \item
    `\texttt{yRange}: A list which details the range of the vertical
    axis.
  \item
    `\texttt{colour}: A list of graphics directives to be applied to
    each data\\
    matrix.
  \end{itemize}
\end{itemize}

There are problems with the default values for x and y ranges and
colour. It would be better to fix the ranges and colour based on the
data as well as the origin of axes and the size of the points.

It would be good if the initial projection matrix (and maybe the centre)
had default values. Perhaps a random matrix with the correct dimensions
read from the data?

The format for the printed projection matrix can be improved also. Maybe
fix the number of digits and the position of the label.

\begin{itemize}
\item
  \texttt{SlicePlot{[}data,\ projmat,\ centrePoint,\ height,\ heightRange,\ ptSize1(=0.005),\ ptSize2(=0.004),\ minDist(=0){]}}
  with arguments (used to be called SliceDynamic in Alex's first
  version, the name SliceDynamic was taken by something else below)

  \begin{itemize}
  \tightlist
  \item
    \texttt{data}: A data matrix
  \item
    \texttt{projmat}: The inital projection matrix
  \item
    \texttt{centrepoint}: The initial position of the slice
  \item
    \texttt{height}: The initial height of the slice
  \item
    \texttt{heightRange}: The range of heights
  \item
    \texttt{ptSize1}: The size of the points that exist within the slice
  \item
    \texttt{ptSize2}: The size of the points that exist outside of the
    slice
  \end{itemize}
\end{itemize}

This function requires some auxiliary functions. It does not accept more
than one data set.

It would be good if the initial projection matrix (and maybe the centre)
had default values. Perhaps a random matrix with the correct dimensions
read from the data?

It may be good to fix the size of the points, the centre of the axes and
the scales of the axes rather than allow them to change as the
projection and slicing changes

\begin{itemize}
\item
  \texttt{SliceDynamic{[}data,\ projmat,\ centrePoint,\ height,\ heightRange,\ ptSize1\ :\ 0.005,\ ptSize2\ :\ 0.004{]}}
  with arguments
\item
  \texttt{data}: A data matrix with an extra groupling or clustering
  column (numerical)
\item
  \texttt{projmat}: The inital projection matrix
\item
  \texttt{centrepoint}: The initial position of the slice
\item
  \texttt{height}: The initial height of the slice
\item
  \texttt{heightRange}: The range of heights
\item
  \texttt{ptSize1}: The size of the points that exist within the slice
\item
  \texttt{ptSize2}: The size of the points that exist outside of the
  slice
\end{itemize}

Does this need to be a separate function from SlicePlot?

It would be good if the initial projection matrix (and maybe the centre)
had default values. Perhaps a random matrix with the correct dimensions
read from the data?

It may be good to fix the size of the points, the centre of the axes and
the scales of the axes rather than allow them to change as the
projection and slicing changes

I have found it useful to input different data sets rather than have
this function sort them automatically using the last column.

\begin{itemize}
\tightlist
\item
  \texttt{Projected2DSliderPlot{[}{]}}
\end{itemize}

This function uses a different display for the projection that I find
very useful. However as it stands it is specific to one data set. Needs
to be generalised to arbitrary data if we want to keep it.

@@

\begin{itemize}
\tightlist
\item
  As was said before, we need to make sure the drawing range is fixed
  (no resizing of the axis when changing projection or slicing).
\item
  Should also make sure the size of the points is fixed (maybe this will
  be automatic if the drawing range is fixed?), bigger points would be
  preferred, currently the size is sometimes really tiny and it becomes
  difficult to see anything.
\item
  Can we have some heuristic to decide the range for the slice thickness
  slider? In the second graph there is a tiny part of the slider where
  the changes are interesting, difficult to navigate since this makes it
  super sensitive to any small movement, while most of the options are
  not interesting. My suggestion would be to keep (as I understand is
  the case now) the maximum thickness (all points in) as the upper
  bound, but instead of zero use a thickness below which slices tend to
  be empty as the lower bound.
\item
  For the olives example I think the relation between x2 and the blue
  region could be interesting.
\item
  pdfsense data: can we use random samples instead of the first 10k
  rows? Then we should still be able to see the bigger picture (with
  picking the first ones we get some artificial structure from how the
  data was sampled initially).
\end{itemize}

I was playing around with this today. Looks like the axes are fixed when
changing slice thickness, but not when rotating?

Anyways, some observations that might be useful, but all of this seems
difficult for including in a paper:

\begin{itemize}
\tightlist
\item
  In the first example we can show how the yellow group is associated
  with the first variable (x1, I guess this is C9?) - when slicing
  through the center it only appears when x1 has a bigger component on
  the projection, it is associated with low values of x1. Since they are
  not near the average value of x1 the points do not get captured in a
  slice where x1 is not important in the projection. We can move the
  center point to -1 in the first component to then explore how the
  yellow group relates to the other three variables. Much less clear,
  maybe some indication of smaller x2 and larger x3 for this.
\item
  The second example is too busy for manual exploration I would say (too
  many axes to move around separately).
\item
  Third example could be similar to the first: blue region requires
  small values of x2, can be understood from the exploration, then move
  center point to -0.5 in x2 to explore further. Now interpretation
  becomes tricky. Sometimes blue and yellow seem just randomly
  everywhere, sometimes blue more contained in an ellipsoid around the
  center (indicating some correlation between variables). For example
  try zeroing out components 2 and 3 to see this slice in x1 vs x4.
\item
  For the last example x4 seems to be the one to work with. I feel like
  this is again very similar, maybe better to instead explore other
  examples (maybe something without slicing?)
\end{itemize}
